Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-09 _tex/paper2.tex
Built site for gh-pages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={Causal analysis, Directed Acyclic Graphs, Bayesian
statistical methods, Thurstonian model, Comparative
judgement, Probability, Statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Causal analysis \sep Directed Acyclic Graphs \sep Bayesian
statistical methods \sep Thurstonian model \sep Comparative
judgement \sep Probability \sep 
    Statistical modeling
\end{keyword}
\end{frontmatter}

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to exhibit a higher trait level. For example, when assessing text
quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity suggests that scores generated by CJ can
accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023},
while research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. The present study primarily focuses on two: the
overreliance on the assumptions of Thurstone's Case V in the statistical
analysis of CJ data, and the apparent disconnect between CJ's approach
to trait measurement and hypothesis testing. The following sections
begin with a brief overview of Thurstone's theory and a detailed
examination of these issues. Subsequently, the study introduces a
theoretical model for CJ that builds upon Thurstone's theory, alongside
its statistical translation, designed to address the two concerns
simultaneously.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons where a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Moreover, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

@@

Building on the previous section, it is clear that, despite its
limitations, the BTL model is commonly used as the measurement model in
CJ assessments. A measurement model specifies how manifest variables
contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
coupled with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can potentially compromise the
reliability of the trait estimates and, by extension, their validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examine the factors influencing CJ
experiments could offer several statistical and measurement benefits,
including the ability to address these issues.

\section{An updated theoretical and statistical model for
CJ}\label{sec-theory}

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal analysis
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014} and, in
particular, directed acyclic graphs (DAGs)
\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
principles of CJ theory. The model also incorporates several practical
factors that influence judges in CJ experiments, such as the selection
of judges, stimuli, and comparisons. In addition, the study uses
Bayesian statistical methods \citep{McElreath_2020} to translate these
theoretical and practical elements into a statistical model that
facilitates the analysis of pairwise comparison data.

@@

@@

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    causal inference \sep directed acyclic graphs \sep structural causal
models \sep bayesian statistical methods \sep thurstonian
model \sep comparative judgement \sep probability \sep 
    statistical modeling
\end{keyword}
\end{frontmatter}

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across different stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to have a higher trait level. For example, when assessing writing
quality, judges compare pairs of written texts (the stimuli) to
determine the relative writing quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on the validity of CJ scores indicates their
capacity to represent the traits under measurement accurately
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Moreover, research on CJ's practical applicability highlights its
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, the
prevalence of unsystematic and fragmented research approaches has left
several critical issues unaddressed. The present study primarily focuses
on two issues: the overreliance on Thurstone's Case V assumptions in the
statistical analysis of CJ data and the apparent disconnect between CJ's
approach to trait measurement and hypothesis testing. The following
sections begin with a brief overview of Thurstone's theory followed by a
detailed examination of these issues. Subsequently, the study introduces
a theoretical model for CJ that builds upon Thurstone's theory,
alongside its statistical translation, designed to address the two
concerns simultaneously.

@@

In its most general form, Thurstone's theory addresses pairwise
comparisons wherein a single judge evaluates multiple stimuli
\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
determine the dichotomous outcome of these comparisons: the discriminal
process of each stimulus and their discriminal difference. The
\emph{discriminal process} captures the psychological impact each
stimulus exerts on the judge or, more simply, his perception of the
stimulus trait. The theory assumes that the discriminal process for any
given stimulus forms a Normal distribution along the trait continuum
\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
known as the \emph{modal discriminal process}, indicates the stimulus
position on this continuum, while its dispersion, referred to as the
\emph{discriminal dispersion}, reflects variability in the perceived
trait of the stimulus.

@@

However, given that the individual discriminal processes of the stimuli
are not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Furthermore, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

@@

Building on the previous section, it is clear that, despite its
limitations, the BTL model is commonly used as a measurement model in CJ
assessments. A measurement model specifies how manifest variables
contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating writing
quality, researchers use the BTL model to process the dichotomous
outcomes resulting from the pairwise comparisons (the manifest
variables) to estimate scores that reflect the underlying level of
writing quality (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
coupled with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can potentially compromise the
reliability of the trait estimates and, by extension, their validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to handling these assumptions and
examining the factors influencing CJ experiments could offer several
statistical and measurement benefits, including the ability to address
these issues.

\section{Updating CJ's theoretical and statistical
model}\label{sec-theory}

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal inference and, in particular, the
structural approach where structural causal models (SCMs) and directed
acyclic graphs (DAGs) articulate the core theoretical principles of CJ
theory. The model also incorporates several assessment design features
that influence CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition, the study uses Bayesian statistical
methods to translate these theoretical and practical elements into a
statistical model that facilitates the analysis of pairwise comparison
data.

@@

\subsubsection{The population model}\label{sec-theory-theoretical_P}

structural causal models (SCMs) and directed acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}

causal analysis \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}

Assuming population data or more commonly known as census data, we
\ldots{}

@@

\subsubsection{The assessment design
features}\label{sec-theory-theoretical_AD}

@@

\section{Appendix}\label{sec-appendix}

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in Section~\ref{sec-theory}. It does not, however, offer a
comprehensive overview of statistical and causal inference methods.
Readers seeking more in-depth understanding may wish to explore
introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
\citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}. They may also find
it helpful to consult introductory books like \citet{Pearl_et_al_2018},
\citet{Neal_2020}, and \citet{McElreath_2020}. For more advanced study,
readers may refer to seminal intermediate papers such as
\citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
\citet{Hernan_et_al_2020}.

\subsection{Empirical research and randomized
experiments}\label{sec-appendix-A}

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where \(E[\cdot]\)
denotes the expected value.

Researchers would then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that transforms data into an estimate \citep{Neal_2020}.
\emph{Estimates} are numerical values that approximate the estimand and
are derived through \emph{estimation}, which refers to the process of
integrating data with an estimator \citep{Everitt_et_al_2010}. The
Identification-Estimation flowchart \citep{McElreath_2020, Neal_2020} in
Figure~\ref{fig-IEflow} provides a visual representation of the process
of transitioning from estimands to estimates.

\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{images/png/IEflow.png}

\caption{\label{fig-IEflow}Identification-Estimation flowchart.
Extracted and slightly modified from \citet[pp.~32]{Neal_2020}}

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable properties that ensure the accuracy
of estimates. For instance, the Z-test is an estimator known for its
effectiveness in comparing groups' proportions, yielding accurate
estimates when the underlying assumptions of the statistic are met
\citep{Kanji_2006}. If this is the case, the Z-test is expressed as a
signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i}| T_{i}=1]\) and \(E[Y_{i}| T_{i}=2]\), respectively. The
noise, represented by \(\hat{s}_{p}\), is defined as the unpooled sample
variability observed between the two groups.

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
identify associations. In the example, researchers can interpret the
associational estimate represented by the Z-statistic as causal. This
interpretation relies on the data meeting the assumptions of the Z-test
and the data being collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers to interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from other variables \(X\) and the outcome \(Y\)
\citep{Morgan_et_al_2014, Neal_2020}.

Nevertheless, researchers often face constraints that limit their
ability to conduct randomized experiments. These constraints include
ethical concerns, such as the assignment of individuals to potentially
harmful interventions, and practical limitations, such as the
infeasibility of, for example, assigning individuals to genetic
modifications or physical impairments \citep{Neal_2020}. In these cases,
causal inference offers a valuable alternative for generating causal
estimates and understanding the mechanisms underlying specific data. In
addition, the framework can provide significant theoretical insights
that can enhance the design of experimental and observational studies
\citep{McElreath_2020}.

\subsection{Empirical research under causal
inference}\label{sec-appendix-B}

Unlike classical statistical modeling, which focuses primarily on
summarizing data and inferring associations, the \emph{causal inference}
framework is designed to identify causes and estimate their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework uses
rigorous mathematical techniques to address the \emph{fundamental
problem of causality}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This problem
revolves around the question, ``What would have happened `in the world'
under different circumstances?'' This question introduces the concept of
counterfactuals, which are instrumental in understanding and defining
causal effects.

\emph{Counterfactuals} represent hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Although a comprehensive
discussion of causes and counterfactuals is beyond the scope of this
document, it is possible to provide a brief overview of how the
framework addresses the fundamental problem of causality. Using the
example outlined in Section~\ref{sec-appendix-B}, the framework begins
by defining the \emph{individual causal effect} (ICE) as the difference
between the students' potential outcomes:
\(\tau_{i} = Y^{1}_{i} - Y^{2}_{i}\). Here, \(Y^{1}_{i}\) represents the
potential outcome observed under \(T_{i}=1\), while \(Y^{2}_{i}\)
represents the potential outcome observed under \(T_{i}=2\). Note that
when a student is assigned to \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) is no longer observed nor observable, making it a
counterfactual. To address the challenge posed by these counterfactuals,
the framework extends the ICE to the \emph{average causal effect} (ACE).
Researchers define the ACE as
\(\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]\), which represents
the difference between the average of observed potential outcomes and
counterfactuals.

Even when data originates from an observational study, researchers can
identify the ACE from associational estimates through causal inference.
They achieve this by performing statistical conditioning on a
\emph{sufficient adjustment set} of variables \(X\)
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially an empty set) should block all (and
not open any) non-causal paths between \(T\) to \(Y\), thereby ensuring
that the estimator for the ACE satisfies several key properties,
including eliminating confounding. If such a set exists, researchers can
express the ACE from associational random quantities as
\(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}. Of course, the validity of the claims about
the effects of \(T\) on \(Y\) now hinges on the assumption that \(X\)
serves as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} noted, ``statements about the real
world from observed (not observational) data require assumptions (and)
this is true no matter what kind of data we have''. For example, if
researchers cannot conduct the randomized experiments described in
Section~\ref{sec-appendix-A} and must rely on observational data, they
can still estimate \(\tau\) if a variable \(X\), such as the
socio-economic status of the school, blocks all (and not open any)
non-causal paths from the teaching method \(T\) to the outcome \(Y\).

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches use rigorous
mathematical notation to characterize the ACE, \emph{not only for
dichotomous but also for continuous variables} \(T\), though they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, as seen in the above examples,
whereas the structural approach uses do-calculus \citep{Pearl_2009}.
Despite these differences, both notations can be expressed in terms of
the other, and both approaches provide methods for using experimental
and observational data to identify and estimate causal effects
\citep{Pearl_2010}.

\subsection{SCMs and DAGs}\label{sec-appendix-C}

\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

The structural approach to causal inference, however, offers a notable
advantage over the potential outcomes approach. It enables researchers
to formally and graphically represent a system using SCMs and DAGs
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. SCMs
and DAGs function as conceptual models describing the presumed causal
structure underlying the ACE. In essence, these models act as
\emph{theoretical models} that guide researchers in determining which
statistical models can yield valid causal inferences, assuming the
depicted causal structure of the models is correct
\citep{McElreath_2020}. Notably, every SCM has a associated DAG
\citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a visual
representation of the role of theoretical models in the inference
process.

SCMs and DAGs offer three key advantages for modeling causal structures.
First, they enable researchers to represent causal relations in a
non-parametric and fully interactive manner. This feature allows for
feasible ACE identification strategies without specifying the data type
or the nature of the functional dependence among variables
\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
represent a wide range of causal structures using only five fundamental
building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures, facilitating their analysis
by focusing on the causal assumptions associated with each building
block \citep{McElreath_2020}. Finally, they help to formalize (causal)
theoretical models into (probabilistic) statistical models for
identification purposes \citep{McElreath_2020} \citep{McElreath_2020}.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks of SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
\(E=\{e_{X},e_{Z},e_{Y}\}\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
are those whose causal mechanisms a researchers chooses to model
\citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

The right panels of the figures display the complementary DAGs. A DAG
consists of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} means the edges extend from one node
to another, with arrows indicating the direction of causal influence.
The term \emph{acyclic} signifies that the causal influences do not form
loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
expected direction of causal influences among these variables. Often,
DAGs omit the exogenous variables for simplicity (the \emph{standard}
representation). However, including exogenous variables in a graph (the
\emph{magnified} representation) can be beneficial, as their presence
can reveal potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

\begin{figure}[H]

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb1}SCM}

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}

\subcaption{\label{fig-mdag_bb1}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-dags_scms1}Two unconnected nodes}

\begin{figure}[H]

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(X,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb2}SCM}

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}

\subcaption{\label{fig-mdag_bb2}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-dags_scms2}Two connected nodes or descendant}

\begin{figure}[H]

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb3}SCM}

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}

\subcaption{\label{fig-mdag_bb3}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-dags_scms3}Chain or mediator}

\begin{figure}[H]

\[
\begin{aligned}
  X & := f_{X}(Z,e_{X}) \\
  Z & := f_{Z}(e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb4}SCM}

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}

\subcaption{\label{fig-mdag_bb4}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-dags_scms4}Fork or confounder}

\begin{figure}[H]

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,Y,e_{Z}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Z} \\
  e_{Z} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_bb5}SCM}

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}

\subcaption{\label{fig-mdag_bb5}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-dags_scms5}Collider or inmorality}

A careful examination of the figures highlights the assumptions
underlying these building blocks. Figures \ref{fig-scm_bb1} and
\ref{fig-mdag_bb1} depict two unconnected nodes, representing a scenario
where variables \(X\) and \(Y\) are not causally related. Figures
\ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two connected nodes,
representing a scenario where a \emph{parent} node \(X\) exerts a causal
influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-scm_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X\) influences \(Z\),
and \(Z\) influences \(Y\). In this configuration, \(X\) is a parent
node of \(Z\), which is a parent node of \(Y\). This structure creates a
\emph{directed path} between \(X\) and \(Y\). Consequently, \(X\) is an
\emph{ancestor} of \(Y\), and \(Z\) fully \emph{mediates} the
relationship between the two. Figures \ref{fig-scm_bb4} and
\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
\(Y\) are both influenced by \(Z\). Here, \(Z\) is a parent node that
\emph{confounds} the relationship between \(X\) and \(Y\). Finally,
Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X\) and \(Y\) are concurrent causes of \(Z\). In this
configuration, \(X\) and \(Y\) are not causally related to each other
but both influence \(Z\) (an \emph{inmorality}). Notably, in all SCMs,
the errors are assumed to be independent of each other and from all
other variables in the graph, as evidenced by the pairwise relations
\(e_{X} \perp\!\!\!\perp e_{Y}\), \(e_{X} \perp\!\!\!\perp e_{Z}\), and
\(e_{Z} \perp\!\!\!\perp e_{Y}\).

Researchers can use these building blocks to depict the scenario
described in Section~\ref{sec-appendix-B}. In this scenario, they rely
on observational data because it is not feasible to run a randomized
experiment. Furthermore, the scenario indicates that researchers have
measured a variable \(X\) (the socio-economic status of the school) that
is assumed to block all non-causal paths from the teaching method \(T\)
to the outcome \(Y\). Figures \ref{fig-scm_example1} and
\ref{fig-mdag_example1} illustrate the plausible causal structure for
this example. The figures highlight the presence of at least four of the
five fundamental building blocks. The figures display multiple
descendants, as shown by pairwise relationships such as
\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
also feature multiple pairs of unconnected nodes, evident from the
relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
Finally, the figures illustrate one fork, \(X \rightarrow \{T, Y\}\),
and two colliders: \(\{X, e_{T}\} \rightarrow T\) and
\(\{X, T, e_{Y}\} \rightarrow Y\).

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \perp\!\!\!\perp e_{X} \\
  e_{T} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_example1}SCM}

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}

\subcaption{\label{fig-mdag_example1}DAG}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\caption{\label{fig-example}Plausible causal structure of the example.}

\subsection{The flow of association and causation in
DAGs}\label{sec-appendix-D}

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{images/png/ACflow.png}

\caption{\label{fig-ACflow}The flow of association and causation in
graphs. Extracted and slightly modified from \citet[pp.~31]{Neal_2020}}
